---
title: "IRF Heterogeneity by Agent Type: State Composition & Individual Income Responses"
author: "Analysis of HTM classification effects on monetary policy transmission"
date: "2026-04-25"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: false
execute:
  warning: false
  message: false
---

# Setup & Data

```{r setup}
library(tidyverse)
library(arrow)
library(lmtest)
library(sandwich)
library(patchwork)

knitr::opts_chunk$set(fig.width = 11, fig.height = 7)

# Set random seed for reproducibility
set.seed(42)

# Create output directories
dir.create("../tables/irf_interaction", showWarnings = FALSE, recursive = TRUE)
dir.create("../tables/irf_individual_income", showWarnings = FALSE, recursive = TRUE)
dir.create("../plots/irf_interaction", showWarnings = FALSE, recursive = TRUE)
dir.create("../plots/irf_individual_income", showWarnings = FALSE, recursive = TRUE)
```

## Load Input Data

```{r load_data}
# State-month panel with shocks and type shares
base <- read_csv("../tables/aggregate_state_monthly_shock_h2m.csv") |>
  filter(!is.na(mp_shock)) |>
  arrange(uf_code, year_month) |>
  group_by(uf_code) |>
  mutate(
    lag_log_c = lag(log(consumption_index + 1)),
    lag_mp_shock = lag(mp_shock)
  ) |>
  ungroup()

# State-quarter HTM shares for representative counterfactuals
htm_shares <- read_csv("../tables/state_quarter_htm_shares.csv") |>
  mutate(uf_code = as.integer(uf_code))

# National average shares by quarter (for counterfactual)
nat_shares <- htm_shares |>
  group_by(year, quarter) |>
  summarise(
    s_ph2m = weighted.mean(share_PH2M, w = total_weight, na.rm = TRUE),
    s_wh2m = weighted.mean(share_WH2M, w = total_weight, na.rm = TRUE),
    s_ric = weighted.mean(share_Ricardian, w = total_weight, na.rm = TRUE),
    .groups = "drop"
  )

# Individual agent types with demographics
ind_types <- read_parquet("../tables/individual_agent_types.parquet")

# PNADC source for income data
pnad_raw <- read_parquet("../../PNAD-C-Treated/pnad_matched.parquet")

cat("Data loaded:\n")
cat("  State-month panel:", nrow(base), "rows\n")
cat("  Individual types:", nrow(ind_types), "rows\n")
cat("  PNADC source:", nrow(pnad_raw), "rows\n")
```

# Helper Functions

## Clustering SE helper

```{r helper_safe_get}
# Extract coefficient and SE safely from coeftest output
safe_get <- function(ct, term) {
  idx <- grep(term, rownames(ct), fixed = FALSE)
  if (length(idx) == 0) {
    return(data.frame(estimate = NA, se = NA, t_stat = NA, pval = NA))
  }
  idx <- idx[1]
  data.frame(
    estimate = ct[idx, 1],
    se = ct[idx, 2],
    t_stat = ct[idx, 3],
    pval = ct[idx, 4]
  )
}

# Compute clustered SEs with optional wild bootstrap
vcov_cluster <- function(model, cluster_var) {
  # cluster_var should be a vector matching nrow(model$model)
  sandwich::vcovCL(model, cluster = cluster_var, type = "HC1")
}

# Extract confidence intervals
get_ci <- function(estimate, se, conf = 0.95) {
  z <- qnorm((1 + conf) / 2)
  list(
    lower = estimate - z * se,
    upper = estimate + z * se
  )
}
```

## Cumulative IRF computation

```{r helper_cumulative_irf}
# Compute cumulative IRF from horizon-specific coefficients
compute_cir <- function(coefs, ses) {
  # coefs: vector of impulse responses [0, 1, 2, ..., h]
  # ses: vector of SEs for each horizon

  h <- length(coefs)
  cir_vals <- cumsum(coefs)

  # SE of cumulative: sqrt(1' Sigma 1) where Sigma is the cov matrix
  # For now, approximate via delta method
  # Exact: need full covariance matrix of all [beta_0, ..., beta_h]
  # Approximation: sqrt(sum(se_i^2)) if assuming independence (conservative)

  cir_ses <- sapply(1:h, function(i) {
    sqrt(sum(ses[1:i]^2))  # conservative: assumes independence
  })

  data.frame(
    horizon = 0:(h-1),
    irf = coefs,
    irf_se = ses,
    cir = cir_vals,
    cir_se = cir_ses
  )
}
```

---

# Approach A: State-Panel Interaction LP

## Run Interaction-based Consumption Response

```{r approach_a_setup}
run_lp_interaction <- function(h, variant = "linear") {
  cat("Running Approach A (h =", h, ")...\n")

  d <- base |>
    filter(!is.na(lag_log_c), !is.na(lag_mp_shock)) |>
    arrange(uf_code, year_month) |>
    group_by(uf_code) |>
    mutate(
      log_c = log(consumption_index + 1),
      y_lead = lead(log_c, h),
      y_resp = y_lead - lag_log_c
    ) |>
    ungroup() |>
    filter(!is.na(y_resp))

  # Extract type shares (using linear variant by default)
  if (variant == "linear") {
    d <- d |>
      mutate(
        share_ph2m = share_PH2M_linear,
        share_wh2m = share_WH2M_linear
      )
  } else {
    d <- d |>
      mutate(
        share_ph2m = share_PH2M_nearest,
        share_wh2m = share_WH2M_nearest
      )
  }

  # Add interaction terms
  d <- d |>
    mutate(
      shock_x_ph2m = mp_shock * share_ph2m,
      shock_x_wh2m = mp_shock * share_wh2m
    )

  # Specification: interaction terms
  fml <- y_resp ~ mp_shock + shock_x_ph2m + shock_x_wh2m +
                  share_ph2m + share_wh2m +
                  lag_log_c + lag_mp_shock +
                  factor(uf_code) + factor(month_num) + factor(year)

  m <- lm(fml, data = d)

  # Cluster SEs by state (27 clusters)
  vcov_c <- vcov_cluster(m, d$uf_code)
  ct <- coeftest(m, vcov. = vcov_c)

  # Extract key coefficients
  beta_mp <- safe_get(ct, "^mp_shock$")
  beta_mp_ph2m <- safe_get(ct, "shock_x_ph2m")
  beta_mp_wh2m <- safe_get(ct, "shock_x_wh2m")

  # Compute implied type-specific IRFs
  # At representative shares: 100% type vs baseline
  irf_ph2m <- beta_mp$estimate + beta_mp_ph2m$estimate  # shock + interaction at share=1
  irf_wh2m <- beta_mp$estimate + beta_mp_wh2m$estimate
  irf_ric <- beta_mp$estimate  # baseline (Ricardian)

  # SEs via delta method (assuming independence of interaction terms)
  se_ph2m <- sqrt(beta_mp$se^2 + beta_mp_ph2m$se^2)
  se_wh2m <- sqrt(beta_mp$se^2 + beta_mp_wh2m$se^2)
  se_ric <- beta_mp$se

  # Return results
  list(
    horizon = h,
    variant = variant,
    model = m,
    vcov = vcov_c,
    coeftest = ct,
    coefficients = data.frame(
      horizon = h,
      beta_mp = beta_mp$estimate,
      beta_mp_se = beta_mp$se,
      beta_mp_ph2m = beta_mp_ph2m$estimate,
      beta_mp_ph2m_se = beta_mp_ph2m$se,
      beta_mp_wh2m = beta_mp_wh2m$estimate,
      beta_mp_wh2m_se = beta_mp_wh2m$se
    ),
    implied_irfs = data.frame(
      horizon = h,
      type = c("PH2M", "WH2M", "Ricardian"),
      irf = c(irf_ph2m, irf_wh2m, irf_ric),
      irf_se = c(se_ph2m, se_wh2m, se_ric)
    ),
    n_obs = nrow(d),
    n_clusters = length(unique(d$uf_code))
  )
}
```

## Run full horizon sweep (h = 0:12)

```{r approach_a_run}
results_a <- map_df(0:12, \(h) {
  res <- run_lp_interaction(h, variant = "linear")
  res$implied_irfs
})

# Add confidence intervals
results_a <- results_a |>
  mutate(
    irf_ci_lower = irf - 1.96 * irf_se,
    irf_ci_upper = irf + 1.96 * irf_se
  )

# Compute cumulative IRF by type
results_a_cir <- results_a |>
  group_by(type) |>
  arrange(horizon) |>
  mutate(
    cir = cumsum(irf),
    cir_se = cumsum(irf_se^2) |> sqrt(),
    cir_ci_lower = cir - 1.96 * cir_se,
    cir_ci_upper = cir + 1.96 * cir_se
  ) |>
  ungroup()

write_csv(results_a, "../tables/irf_interaction/implied_type_irfs_linear.csv")
write_csv(results_a_cir, "../tables/irf_interaction/implied_type_irfs_cir_linear.csv")

cat("Approach A results saved:\n")
print(head(results_a, 10))
```

---

# Approach B: Individual Income Response LP

## Construct Individual Panel with Income

```{r approach_b_prep}
cat("Constructing individual income panel...\n")

# Merge individual types with PNADC income
# First filter PNADC to have id_ind and select needed columns
pnad_income <- pnad_raw |>
  filter(!is.na(id_ind)) |>
  select(id_ind, Ano, Trimestre, rendimento_habitual_real) |>
  distinct()

ind_income <- ind_types |>
  select(id_ind, id_dom, year, quarter, uf_code, weight, agent_type) |>
  left_join(
    pnad_income,
    by = c("id_ind", "year" = "Ano", "quarter" = "Trimestre")
  )

# Diagnostic: check merge quality
merge_diag <- ind_income |>
  summarise(
    n_total = n(),
    n_has_income = sum(!is.na(rendimento_habitual_real)),
    pct_missing = 100 * mean(is.na(rendimento_habitual_real))
  )

print(merge_diag)
write_csv(merge_diag, "../diagnostics/individual_income_merge.csv")

# Filter to records with income; add log transformation
ind_income <- ind_income |>
  filter(!is.na(rendimento_habitual_real)) |>
  mutate(
    log_inc = log(rendimento_habitual_real + 1)
  )

cat("Individual income panel:", nrow(ind_income), "rows\n")
```

## Merge quarterly monetary shocks

```{r approach_b_shocks}
# Aggregate monthly shocks to quarterly
shock_qtr <- base |>
  mutate(quarter = ceiling(month_num / 3)) |>
  group_by(uf_code, year, quarter) |>
  summarise(
    mp_shock_qtr = mean(mp_shock, na.rm = TRUE),
    n_months = n(),
    .groups = "drop"
  )

# Merge shocks onto individual panel
ind_income <- ind_income |>
  left_join(
    shock_qtr,
    by = c("uf_code", "year", "quarter")
  )

# Check merge quality
shock_merge_diag <- ind_income |>
  summarise(
    n_total = n(),
    n_has_shock = sum(!is.na(mp_shock_qtr)),
    pct_unmatched = 100 * mean(is.na(mp_shock_qtr))
  )

print(shock_merge_diag)

# Filter to matched records
ind_income <- ind_income |>
  filter(!is.na(mp_shock_qtr))

cat("Individual panel with shocks:", nrow(ind_income), "rows\n")
```

## Individual LP by type

```{r approach_b_lp}
run_lp_individual <- function(h, type_filter, data) {
  cat("Running individual LP (h =", h, ", type =", type_filter, ")...\n")

  d <- data |>
    filter(agent_type == type_filter) |>
    group_by(id_ind) |>
    arrange(year, quarter) |>
    mutate(
      log_inc_lag = lag(log_inc, 1),
      y_lead = lead(log_inc, h),
      y_resp = y_lead - log_inc_lag,
      shock_lag = lag(mp_shock_qtr, 1)
    ) |>
    ungroup() |>
    filter(!is.na(y_resp), !is.na(log_inc_lag), !is.na(shock_lag))

  # Check factor levels - only include if >1 level
  formula_parts <- c("y_resp ~ mp_shock_qtr + log_inc_lag + shock_lag")
  if (n_distinct(d$uf_code) > 1) formula_parts <- c(formula_parts, "+ factor(uf_code)")
  if (n_distinct(d$year) > 1) formula_parts <- c(formula_parts, "+ factor(year)")
  if (n_distinct(d$quarter) > 1) formula_parts <- c(formula_parts, "+ factor(quarter)")

  fml <- as.formula(paste(formula_parts, collapse = " "))

  m <- lm(fml, data = d, weights = d$weight)

  # Cluster SEs by state
  vcov_c <- vcov_cluster(m, d$uf_code)
  ct <- coeftest(m, vcov. = vcov_c)

  # Extract main shock coefficient
  coef_shock <- safe_get(ct, "^mp_shock_qtr$")

  list(
    horizon = h,
    type = type_filter,
    model = m,
    coef_shock = coef_shock,
    n_obs = nrow(d),
    n_unique_ind = n_distinct(d$id_ind),
    n_clusters = length(unique(d$uf_code))
  )
}

# Run sweep across all horizons and types
results_b <- expand_grid(
  h = 0:8,
  type = c("PH2M", "WH2M", "Ricardian")
) |>
  mutate(
    result = map2(h, type, \(h, t) run_lp_individual(h, t, ind_income))
  ) |>
  unnest_wider(result) |>
  select(horizon, type, coef_shock) |>
  unnest_wider(coef_shock) |>
  rename(
    irf = estimate,
    irf_se = se,
    irf_pval = pval
  )

write_csv(results_b, "../tables/irf_individual_income/irfs_by_type.csv")

cat("Approach B results saved:\n")
print(head(results_b, 12))
```

## Cumulative IRF for individual approach

```{r approach_b_cir}
results_b_cir <- results_b |>
  group_by(type) |>
  arrange(horizon) |>
  mutate(
    cir = cumsum(irf),
    cir_se = cumsum(irf_se^2) |> sqrt(),
    cir_ci_lower = cir - 1.96 * cir_se,
    cir_ci_upper = cir + 1.96 * cir_se,
    irf_ci_lower = irf - 1.96 * irf_se,
    irf_ci_upper = irf + 1.96 * irf_se
  ) |>
  ungroup()

write_csv(results_b_cir, "../tables/irf_individual_income/irfs_cir_by_type.csv")

cat("Individual income cumulative IRFs:\n")
print(head(results_b_cir, 12))
```

---

# Master Results Synthesis

## Combine both approaches

```{r master_synthesis}
# Stack Approach A and B into master table
master_a <- results_a_cir |>
  rename(
    impulse_estimate = irf,
    impulse_se = irf_se,
    impulse_ci_lower = irf_ci_lower,
    impulse_ci_upper = irf_ci_upper,
    cir_estimate = cir,
    cir_se = cir_se,
    cir_ci_lower = cir_ci_lower,
    cir_ci_upper = cir_ci_upper
  ) |>
  mutate(
    approach = "state_interaction",
    outcome = "consumption",
    frequency = "monthly",
    se_method = "clustered",
    n_obs = NA,
    n_clusters = NA,
    pvalue = NA
  ) |>
  select(approach, outcome, frequency, type, horizon,
         impulse_estimate, impulse_se, impulse_ci_lower, impulse_ci_upper, pvalue,
         cir_estimate, cir_se, cir_ci_lower, cir_ci_upper,
         n_obs, n_clusters, se_method)

master_b <- results_b_cir |>
  rename(
    impulse_estimate = irf,
    impulse_se = irf_se,
    impulse_ci_lower = irf_ci_lower,
    impulse_ci_upper = irf_ci_upper,
    cir_estimate = cir,
    cir_se = cir_se,
    cir_ci_lower = cir_ci_lower,
    cir_ci_upper = cir_ci_upper,
    pvalue = irf_pval,
    agent_type = type
  ) |>
  rename(type = agent_type) |>
  mutate(
    approach = "individual_income",
    outcome = "income",
    frequency = "quarterly",
    se_method = "clustered",
    n_obs = NA,
    n_clusters = NA
  ) |>
  select(approach, outcome, frequency, type, horizon,
         impulse_estimate, impulse_se, impulse_ci_lower, impulse_ci_upper, pvalue,
         cir_estimate, cir_se, cir_ci_lower, cir_ci_upper,
         n_obs, n_clusters, se_method)

master_results <- bind_rows(master_a, master_b)

write_parquet(master_results, "../tables/irf_master_results.parquet")

cat("Master results table:\n")
print(head(master_results, 20))
cat("Total rows:", nrow(master_results), "\n")
```

---

# Visualizations

## Approach A: Interaction-based consumption IRFs

```{r plot_approach_a}
p_irfs_a <- results_a_cir |>
  ggplot(aes(x = horizon, y = irf, color = type, fill = type)) +
  geom_ribbon(aes(ymin = irf_ci_lower, ymax = irf_ci_upper), alpha = 0.2, color = NA) +
  geom_line(size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Approach A: Consumption IRF by Type (State-Panel Interactions)",
    x = "Horizon (months)",
    y = "Impulse Response",
    color = "Agent Type",
    fill = "Agent Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

p_cir_a <- results_a_cir |>
  ggplot(aes(x = horizon, y = cir, color = type, fill = type)) +
  geom_ribbon(aes(ymin = cir_ci_lower, ymax = cir_ci_upper), alpha = 0.2, color = NA) +
  geom_line(size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Approach A: Cumulative Consumption IRF by Type",
    x = "Horizon (months)",
    y = "Cumulative IRF",
    color = "Agent Type",
    fill = "Agent Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave("../plots/irf_interaction/irfs.png", p_irfs_a, width = 10, height = 6, dpi = 300)
ggsave("../plots/irf_interaction/cumulative_irfs.png", p_cir_a, width = 10, height = 6, dpi = 300)

p_irfs_a
p_cir_a
```

## Approach B: Individual income IRFs

```{r plot_approach_b}
p_irfs_b <- results_b_cir |>
  ggplot(aes(x = horizon, y = irf, color = type, fill = type)) +
  geom_ribbon(aes(ymin = irf_ci_lower, ymax = irf_ci_upper), alpha = 0.2, color = NA) +
  geom_line(size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Approach B: Income IRF by Type (Individual LP)",
    x = "Horizon (quarters)",
    y = "Impulse Response",
    color = "Agent Type",
    fill = "Agent Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

p_cir_b <- results_b_cir |>
  ggplot(aes(x = horizon, y = cir, color = type, fill = type)) +
  geom_ribbon(aes(ymin = cir_ci_lower, ymax = cir_ci_upper), alpha = 0.2, color = NA) +
  geom_line(size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Approach B: Cumulative Income IRF by Type",
    x = "Horizon (quarters)",
    y = "Cumulative IRF",
    color = "Agent Type",
    fill = "Agent Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave("../plots/irf_individual_income/irfs.png", p_irfs_b, width = 10, height = 6, dpi = 300)
ggsave("../plots/irf_individual_income/cumulative_irfs.png", p_cir_b, width = 10, height = 6, dpi = 300)

p_irfs_b
p_cir_b
```

## Cross-approach synthesis plot

```{r plot_synthesis}
# Resample Approach A to quarterly granularity for comparison
# (every 3 months)
results_a_qtr <- results_a_cir |>
  filter(horizon %% 3 == 0) |>
  mutate(horizon_qtr = horizon / 3)

# Plot all 6 panels: rows = approach, cols = outcome type
p_synthesis <- expand_grid(
  outcome = c("Approach A: Consumption", "Approach B: Income"),
  type = c("PH2M", "WH2M", "Ricardian")
) |>
  mutate(
    data = case_when(
      grepl("A", outcome) ~ list(results_a_cir |> filter(horizon <= 12) |> mutate(approach = "A")),
      .default ~ list(results_b_cir |> mutate(approach = "B"))
    ),
    p = pmap(list(outcome, type, data), function(out, t, d) {
      d_filt <- d |> filter(type == t)
      ggplot(d_filt, aes(x = horizon, y = irf)) +
        geom_ribbon(aes(ymin = irf_ci_lower, ymax = irf_ci_upper),
                   fill = "lightblue", alpha = 0.5) +
        geom_line(color = "steelblue", size = 1) +
        geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
        labs(title = paste(out, "-", t),
             x = "Horizon", y = "IRF") +
        theme_minimal() +
        theme(plot.title = element_text(size = 10))
    })
  ) |>
  pull(p) |>
  wrap_plots(ncol = 3)

ggsave("../plots/irf_synthesis_comparison.png", p_synthesis, width = 14, height = 8, dpi = 300)

p_synthesis
```

---

# Summary Statistics

```{r summary}
cat("\n=== APPROACH A: STATE-INTERACTION CONSUMPTION LP ===\n")
cat("Horizons: 0-12 months\n")
cat("Outcome: Log consumption (state-level index)\n")
cat("Specification: y_resp ~ mp_shock + mp_shock×share_PH2M + mp_shock×share_WH2M + controls + FE\n")
cat("Type-specific IRF = β_shock + β_shock×type × 1 (at 100% type share)\n\n")

cat("Implied consumption responses at h=0 (impact):\n")
print(results_a_cir |> filter(horizon == 0))

cat("\n=== APPROACH B: INDIVIDUAL INCOME LP ===\n")
cat("Horizons: 0-8 quarters\n")
cat("Outcome: Log real habitual income (individual)\n")
cat("Specification: y_resp ~ mp_shock + controls + state×quarter FE, stratified by type\n\n")

cat("Income responses at h=0 (impact):\n")
print(results_b_cir |> filter(horizon == 0))

cat("\n=== CROSS-APPROACH DIRECTIONAL CONSISTENCY CHECK ===\n")
# Check if type orderings match between approaches
a_order <- results_a_cir |>
  filter(horizon == 0) |>
  arrange(desc(irf)) |>
  pull(type)

b_order <- results_b_cir |>
  filter(horizon == 0) |>
  arrange(desc(irf)) |>
  pull(type)

cat("Approach A type ranking (highest to lowest h=0 response):", paste(a_order, collapse=" > "), "\n")
cat("Approach B type ranking (highest to lowest h=0 response):", paste(b_order, collapse=" > "), "\n")
cat("Directional consistency:", ifelse(identical(a_order, b_order), "✓ YES", "✗ NO"), "\n")
```

